The effects of binarity on planet occurrence rates measured by transit surveys

### 2018-02-10 paper/main.tex
clarifications based on KM suggestions

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{figures/visualize_volumes.pdf}
    \end{center}
    \caption{
        Searchable volumes for single stars ({\it left}) and twin
        binaries ({\it right}), assuming that all stars have the same
        mass, radius, and luminosity, and that all planets have the
        same radius $r=r\p$.
        {\it Top:} At an apparent radius $r\a=r\p$, the observer
        searched twin binaries out to a distance $\sqrt{2}\times$ that
        of single stars.  Because of dilution, there are no planets in
        twin binaries with $r\a=r\p$.
        {\it Middle:} At $r\a=r\p/\sqrt{2}$, the maximum searchable
        distances are half those at $r\a=r\p$.  The only detected
        planets with $r\a=r\p/\sqrt{2}$ orbit twin binaries.
        {\it Bottom:} Planets with a true radius $r=r\p$ are
        searchable to a maximum distance $d_0$ around singles, and
        $d_0/\sqrt{2}$ around twin binaries.
    }
    \label{fig:model_1_volumes}
\end{figure}

@@

Let us evaluate the contributions to $n_{\rm det}$ from singles and
binaries.  As in the previous subsection, for each apparent planet
radius $r\a$, the observers select $N_\star(r\a)$ unresolved
point-sources on sky, based on Equation \ref{eq:dmax}.  If we let $\mu
\equiv N\star^{\rm b}(r\a)/N_\star^0(r\a)$ be the ratio of the number
of searchable binary systems to single systems, then the number of
single and binary point-sources thought to be searchable are
\begin{equation}
  \frac{N_\star(r\a)}{1+\mu}~~{\rm and}~~\frac{\mu N_\star(r\a)}{1+\mu},
\end{equation}
respectively.  Earlier we showed that due to Malmquist bias, $\mu =
2^{3/2}(n\b/n\s)$.

@@

We will now apply our general equation for the apparent rate density
(Equation~\ref{eq:general_Gamma_a}) to study binarity's effects in
regimes of observational interest.  In the following, we write the
rate density for each type of star, $\Gamma_i(r)$, as the product of a
shape function and a constant:
\begin{equation}
    \Gamma_i(r) = Z_i f_i(r), \quad {\rm for\ }i\in\{0,1,2\},
\end{equation}
where $i=0$ corresponds to single stars, $i=1$ to primaries of
binaries, and $i=2$ to secondaries of binaries.  The shape function is
normalized to unity.  The $Z_i$'s are each system type's occurrence
rate $\Lambda_i$, integrated over all planetary radii. In other words,
they are number of planets per single, primary, or secondary star.  We
want to consider the effects of varying not only $f_i(r)$ but also the
relative values of the $Z_i$'s.

@@

\subsubsection{Twin binaries}
We begin introducing realism by keeping all binaries as twins, but
letting the planet radius distributions vary.  First, we assume a
power law planet radius distribution,
\begin{equation}
    \Gamma_i(r) = Z_i f(r) = Z_i r^\delta/\mathcal{N}_r,
\end{equation}
for $f(r)$ the radius shape function, and $\mathcal{N}_r$ the shape
function's normalization.  The resulting apparent rate density is 
\begin{equation}
    \Gamma\a(r\a) = \frac{r\a^\delta}{\mathcal{N}_r} \left[
    \frac{Z_0}{1+\mu}
    +
    2^{\frac{\delta+1}{2} } \frac{\mu}{1+\mu} \left(Z_1 + Z_2
    \right)
    \right],
    \label{eq:model5_apparent_rate_density}
\end{equation}
which is quite similar to the twin binary, fixed-planet case
(Equation~\ref{eq:model_1_apparent_rate_density}), except for in the
radius dependence and normalization.  As in Section~\ref{sec:model_1},
$\mu = 2^{3/2} n\b/n\s$.  If the $Z_i$'s are equal, the ``correction
factor'' relative to the rate density for singles becomes quite
simple:
\begin{align}
    \left. \frac{\Gamma\a(r\a)}{\Gamma_0(r)} 
    \right|_{r\a\rightarrow r}
    &=
    \frac{1 + 2^{\frac{\delta+3}{2}}\mu}{1 + \mu}.
    \label{eq:power_law_correction}
\end{align}
If the binary fraction is $0.1$, $\mu\approx 0.15$. Taking
$\delta=-2.92$ from \citet{howard_planet_2012},  we find
$\Gamma\a/\Gamma_0 = 1.004$.  In other words, the apparent rate
density is an {\it over}estimate compared to the rate density of
single stars, with a relative difference $\delta \Gamma_0 =
|\Gamma_0~-~\Gamma\a|/\Gamma_0$ of 0.4\%.  This is quite a small
effect!  Clearly though, it cannot apply at small radii, since $f(r)
\propto r^\delta$ diverges; \citet{howard_planet_2012}'s results
indicate that the power law planet radius distribution holds for
$r\a\gtrsim 2r_\oplus$.  It is similarly nonsensical for $f(r)$ to be
finite above some upper radius limit $r_{\rm u}$, perhaps around
$\approx 24r_\oplus$, based on the most inflated hot Jupiters.
Imposing an upper cut-off of $r_{\rm u}$ would lead to smaller values
of $\Gamma(r\a)$ down to $r_{\rm u}/\sqrt{2}$, compared to an
intrinsic power law without the cut-off.  We are not particularly
interested in this effect because there are better models for hot
Jupiter occurrence rates that we will discuss in
Section~\ref{sec:further_models}.  If the assumptions behind this
model are at all applicable to real transit surveys,
Equation~\ref{eq:power_law_correction} suggests that for $2r_\oplus
\lesssim r\a \lesssim 17r_\oplus$, binarity's impact on apparent
occurrence rates is quite small.

@@

To check whether non-twin binaries change the preceding result, we now
assume $f(q) = q^\beta/\mathcal{N}_q$, for $\mathcal{N}_q$ the
normalization.  This changes $\mu$ (Equation~\ref{eq:mu_general}
simplifies to Equation~\ref{eq:mu_power_laws}).  It may also affect
the rate density, which could be a function of the varying stellar
mass.  Absorbing this dependence into a power law as well,
\begin{equation}
    \Gamma_i(r,M) = Z_i \times \frac{r^\delta}{\mathcal{N}_r} \times
    \frac{M^\gamma}{\mathcal{N}_M},
\end{equation}
where $Z_i$ is dimensionless, and the normalization constants carry
the units (each side has units $[r^{-1} M^{-1}]$).  We assume that
stars are a one-parameter family, given by $R \propto M \propto
L^{\frac{1}{\alpha}}$, so that a given value of $q$ determines
everything about a secondary.

We are mostly interested in the apparent rate density's radius
dependence.  Marginalizing Equation~\ref{eq:general_Gamma_a} over
apparent stellar mass, we find that when $Z_0=Z_1=Z_2$,
\begin{align}
    X_\Gamma &\equiv \left. \frac{\Gamma\a(r\a)}{\Gamma_0(r)} 
    \right|_{r\a\rightarrow r} \\
    &=
    \notag
    \frac{1}{1+\mu}
    \left[1 + \frac{1}{\mathcal{N}_q} \frac{n\b}{n\s}\times 
    \left(
    \int_0^1 {\rm d}q\,q^\beta (1+q^\alpha)^{\frac{\delta+4}{2}} +
    \right.
    \right. \\
    %isn't latex great?
    &\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad
    \left.\left.
    \int_0^1 {\rm d}q\,q^{\beta+\delta+\frac{5}{3}} 
    (1+q^\alpha)^{\frac{3}{2}}
    (1+q^{-\alpha})^{\frac{\delta+1}{2}}
    \right)\right],
    \label{eq:powerlaw_vary_binary}
\end{align}
where $\gamma$ does not appear because of the marginalization over
$M\a$.  For $\alpha = 3.5$, $\beta=0$, $\gamma=0$, $\delta=-2.92$, the
summed integrals in Equation~\ref{eq:powerlaw_vary_binary} give
$(\ldots)\approx 1.50249$. %171211_model_integrals.nb
For $n\b / (n\b + n\s)=0.44$, this yields $\Gamma\a/\Gamma_0 = 1.048$;
the relative difference between the apparent rate density and the rate
density around singles is 4.8\%.  This indicates that considering only
twin binaries gave us correct intuition: for a power law radius
distribution ($2r_\oplus \lesssim r\a \lesssim 17r_\oplus$) in which
there are the same number of planets orbiting singles, primaries, and
secondaries, binarity influences apparent planet occurrence rates
around Sun-like stars at the $\sim$few percent level.

@@

Though the details of the planet radius distribution $\Gamma_i(r)$ at
$r<2r_\oplus$ are currently an active topic of research, we can
consider how binarity affects this regime under different plausible
assumptions.  For instance, assume that the true radius shape function
is
\begin{align}
    f(r)
    &\propto
    \left.
    \begin{cases}
        r^\delta & \text{for } r\geq 2r_\oplus \\
        {\rm constant} & \text{for } r\leq2r_\oplus.
    \end{cases}
    \right.
    \label{eq:model3_radius_distribution}
\end{align}
As in the previous model, we presume that the observers know the true
properties of all singles and primaries. The masses, radii, and
luminosities of stars vary as $R \propto M \propto
L^{\frac{1}{\alpha}}$.  Our ``nominal model'' remains the same:
$\alpha=3.5$, $\beta=\gamma=0$, $\delta=-2.92$.

At apparent radii $r\a > 2r_\oplus$, the results of this model are the
same as those from Section~\ref{sub:powerlaw_varying_binaries}.  For
$r\a < 2r_\oplus$, the equations are tedious, but still tractable.
For simplicity, we insert Equation~\ref{eq:model3_radius_distribution}
into Equation~\ref{eq:general_Gamma_a}, and integrate using a computer
program. We refer the interested reader to our online
implementation\footnote{\url{https://github.com/lgbouma/binary_biases}}.
The output is validated against analytic predictions in the $r\a >
2r_\oplus$ and the $r\a < 2r_\oplus/\sqrt{2}$ regimes, and is plotted
in Figure~\ref{fig:occ_rate_model_3_log}.

\paragraph{The rate of Earth analogs} The immediately arresting result
there is a ``bump'' in the apparent rate density at $r\a < 2r_\oplus$:
the true rate for singles is less than the inferred rate.  For the
case in which secondaries host as many (half as many) planets as
single stars, this means an overestimate of the absolute occurrence
rate by $\approx 50\%$ ($\approx 25\%$).  The ``bump'' exists even for
the $Z_2/Z_0=0$ case as a $\approx 10\%$ effect.  The magnitude of
this evidently error depends strongly on the prevalence of planets
around secondaries.

\paragraph{Hot Jupiter occurrence rates} Taking
Figure~\ref{fig:occ_rate_model_3_log} and integrating the rate
density, we can compare the apparent hot Jupiter occurrence rate with
the true rate for singles.  In the most extreme case of $Z_2/Z_0=0$,
we find that $\Lambda_{{\rm HJ},0}/\Lambda_{\rm HJ,a} = 1.13$, where
\begin{equation}
    \Lambda_{\rm HJ,a} =
      \int_{8r_\oplus}^{\infty} \Gamma\a(r)\,{\rm d}r,
\end{equation}
and similarly for $\Lambda_{\rm HJ,0}$.
%occ_rate_vs_radius_model_3_withtext_Zsub2_0.00_rpu_22.5.pdf
If $Z_2/Z_0>0$, binarity affects the apparent hot Jupiter rate less:
when $Z_2/Z_0=0.5$, we find $\Lambda_{{\rm HJ},0}/\Lambda_{\rm HJ,a} =
1.06$.  We discuss this issue further in
Section~\ref{sec:further_models}.

>>>>

\begin{figure}[!tb]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{figures/visualize_volumes.pdf}
    \end{center}
    \caption{
        Searchable volumes for single stars ({\it left}) and twin
        binaries ({\it right}), assuming that all stars have the same
        mass, radius, and luminosity, and that all planets have the
        same radius $r=r\p$.
        {\it Top:} At an apparent radius $r\a=r\p$, the observer
        searched twin binaries out to a distance $\sqrt{2}\times$ that
        of single stars.  Because of dilution, there are no planets in
        twin binaries with $r\a=r\p$.
        {\it Bottom:} At $r\a=r\p/\sqrt{2}$, the maximum searchable
        distances are half those at $r\a=r\p$.  The only detected
        planets with $r\a=r\p/\sqrt{2}$ orbit twin binaries.
    }
    \label{fig:model_1_volumes}
\end{figure}

@@

Let us evaluate the contributions to $n_{\rm det}$ from singles and
binaries.  As in the previous subsection, for each apparent planet
radius $r\a$, the observers select $N_\star(r\a)$ unresolved
point-sources on sky, based on Equation \ref{eq:dmax}.  If we let $\mu
\equiv N_\star^{\rm b}(r\a)/N_\star^0(r\a)$ be the ratio of the number
of searchable binary systems to single systems, then the number of
single and binary point-sources thought to be searchable are
\begin{equation}
  \frac{N_\star(r\a)}{1+\mu}~~{\rm and}~~\frac{\mu N_\star(r\a)}{1+\mu},
\end{equation}
respectively.  Earlier we showed that due to Malmquist bias, $\mu =
2^{3/2}(n\b/n\s)$.

@@

We will now apply our general equation for the apparent rate density
(Equation~\ref{eq:general_Gamma_a}) to study binarity's effects in
regimes of observational interest.  In the following, we write the
rate density for each type of star, $\Gamma_i(r)$, as the product of a
shape function, normalized to unity, and a constant:
\begin{equation}
    \Gamma_i(r) = Z_i f_i(r), \quad {\rm for\ }i\in\{0,1,2\},
\end{equation}
where $i=0$ corresponds to single stars, $i=1$ to primaries of
binaries, and $i=2$ to secondaries of binaries.
The $Z_i$'s are each system type's occurrence
rate $\Lambda_i$, integrated over all planetary radii. In other words,
they are number of planets per single, primary, or secondary star.
While our main interest is in varying the radius shape function
and the relative values of integrated occurrence rates, we 
will also briefly discuss shape functions in stellar mass.
To keep things analytically simple, we will assume throughout this
section that stars are a one-parameter family, given by 
$L \propto M^\alpha \propto R^\alpha$, so that a given value 
of $q$ determines everything about a secondary.

@@

\subsubsection{Twin binaries}
As a first step, we keep all binaries as twins, but
let the planet radius distribution vary.  If we take a
power law planet radius distribution,
\begin{equation}
    \Gamma_i(r) = Z_i f(r) = Z_i r^\delta/\mathcal{N}_r,
\end{equation}
for $\mathcal{N}_r$ the shape function's normalization, then
Equation~\ref{eq:general_Gamma_a} gives the resulting apparent rate density,
\begin{equation}
    \Gamma\a(r\a) = \frac{r\a^\delta}{\mathcal{N}_r} \left[
    \frac{Z_0}{1+\mu}
    +
    2^{\frac{\delta+1}{2} } \frac{\mu}{1+\mu} \left(Z_1 + Z_2
    \right)
    \right].
    \label{eq:model5_apparent_rate_density}
\end{equation}
This is similar to the twin binary, fixed-planet case
(Equation~\ref{eq:model_1_apparent_rate_density}), except for the
radius dependence and normalization.  As in Section~\ref{sec:model_1},
$\mu = 2^{3/2} n\b/n\s$.  If the $Z_i$'s are equal, the ``correction
factor'' relative to the rate density for singles becomes simple:
\begin{align}
    \left. \frac{\Gamma\a(r\a)}{\Gamma_0(r)} 
    \right|_{r\a\rightarrow r}
    &=
    \frac{1 + 2^{\frac{\delta+3}{2}}\mu}{1 + \mu}.
    \label{eq:power_law_correction}
\end{align}
If the binary fraction is $0.1$, $\mu\approx 0.15$. Taking
$\delta=-2.92$ from \citet{howard_planet_2012},  we find
$\Gamma\a/\Gamma_0 = 1.004$.  In other words, the apparent rate
density is an {\it over}estimate compared to the rate density of
single stars, with a relative difference $\delta \Gamma_0 =
|\Gamma_0~-~\Gamma\a|/\Gamma_0$ of 0.4\%.  This is quite a small
effect!  Clearly though, it cannot apply at small radii, since $f(r)
\propto r^\delta$ diverges; \citet{howard_planet_2012}'s results
indicate that the power law planet radius distribution holds for
$r\a\gtrsim 2r_\oplus$.  It is similarly nonsensical for $f(r)$ to be
finite above some upper radius limit $r_{\rm u}$, perhaps around
$\approx 24r_\oplus$, based on the most inflated hot Jupiters.
Imposing an upper cut-off of $r_{\rm u}$ would lead to smaller values
of $\Gamma(r\a)$ down to $r_{\rm u}/\sqrt{2}$, compared to an
intrinsic power law without the cut-off.  We are not particularly
interested in this effect because there are better models for hot
Jupiter occurrence rates that we will discuss in
Section~\ref{sec:further_models}.  If the assumptions behind this
model are at all applicable to real transit surveys,
Equation~\ref{eq:power_law_correction} suggests that for $2r_\oplus
\lesssim r\a \lesssim 17r_\oplus$, binarity's impact on apparent
occurrence rates is quite small.

@@

To check whether non-twin binaries change the preceding result, we now
assume $f(q) = q^\beta/\mathcal{N}_q$, for $\mathcal{N}_q$ the
normalization.
This changes $\mu$: taking the volume-limited binary mass ratio
distribution to be $f(q) \propto q^\beta$, Equation~\ref{eq:mu_general} 
simplifies to
\begin{equation}
    \mu = \frac{n\b}{n\s} \frac{1}{1+\beta}
    \int_0^1 (1+q^\alpha)^{3/2}q^\beta\,{\rm d}q.
    \label{eq:mu_power_law_q}
\end{equation}
 The rate density might now also vary with 
host star mass.  We can absorb this dependence into a power
law,
\begin{equation}
    \Gamma_i(r,M) = Z_i \times \frac{r^\delta}{\mathcal{N}_r} \times
    \frac{M^\gamma}{\mathcal{N}_M},
\end{equation}
where $Z_i$ is dimensionless, and the normalization constants carry
the units (each side has units $[r^{-1} M^{-1}]$).

We are mostly interested in the apparent rate density's radius
dependence.  Marginalizing Equation~\ref{eq:general_Gamma_a} over
apparent stellar mass, we find that when all stars host the same number
of planets ($Z_0=Z_1=Z_2$),
\begin{align}
    \notag
    \left. \frac{\Gamma\a(r\a)}{\Gamma_0(r)} 
    \right|_{r\a\rightarrow r}
    &=
    \frac{1}{1+\mu}
    \left[1 + \frac{1}{\mathcal{N}_q} \frac{n\b}{n\s}\times 
    \left(
    \int_0^1 {\rm d}q\,q^\beta (1+q^\alpha)^{\frac{\delta+4}{2}} +
    \right.
    \right. \\
    %isn't latex great?
    &\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad
    \left.\left.
    \int_0^1 {\rm d}q\,q^{\beta+\delta+\frac{5}{3}} 
    (1+q^\alpha)^{\frac{3}{2}}
    (1+q^{-\alpha})^{\frac{\delta+1}{2}}
    \right)\right],
    \label{eq:powerlaw_vary_binary}
\end{align}
where $\gamma$ does not appear because of the marginalization over
$M\a$.  For $\alpha = 3.5$, $\beta=0$, $\gamma=0$, $\delta=-2.92$, the
summed integrals in Equation~\ref{eq:powerlaw_vary_binary} give
$(\ldots)\approx 1.503$. %171211_model_integrals.nb
For $n\b / (n\b + n\s)=0.44$, this yields $\Gamma\a/\Gamma_0 = 1.048$;
the relative difference between the apparent rate density and the rate
density around singles is 4.8\%.  This indicates that considering only
twin binaries gave us correct intuition: for a power law radius
distribution ($2r_\oplus \lesssim r\a \lesssim 17r_\oplus$) in which
there are the same number of planets orbiting singles, primaries, and
secondaries, binarity influences apparent planet occurrence rates
around Sun-like stars at the few percent level.

@@

Though the details of the planet radius distribution $\Gamma_i(r)$ at
$r<2r_\oplus$ are currently an active topic of research, we can
investigate how binarity affects this regime by making plausible
assumptions.  For instance, we can assume a broken power-law:
\begin{align}
    f(r)
    &\propto
    \left.
    \begin{cases}
        r^\delta & \text{for } r\geq 2r_\oplus \\
        {\rm constant} & \text{for } r\leq2r_\oplus.
    \end{cases}
    \right.
    \label{eq:model3_radius_distribution}
\end{align}
Our ``nominal model'' remains the same:
$L\propto M^\alpha \propto R^\alpha$, and
$\alpha=3.5$, $\beta=\gamma=0$, $\delta=-2.92$.

At apparent radii $r\a > 2r_\oplus$, the apparent rate density in
this model is the same as from
Section~\ref{sub:powerlaw_varying_binaries}.  For
$r\a < 2r_\oplus$, the equations are tedious, but still tractable.
For simplicity, we insert Equation~\ref{eq:model3_radius_distribution}
into Equation~\ref{eq:general_Gamma_a}, and integrate using a computer
program. We refer the interested reader to our online
implementation\footnote{\url{https://github.com/lgbouma/binary_biases}}.
The output is validated against analytic predictions in the $r\a >
2r_\oplus$ and the $r\a < 2r_\oplus/\sqrt{2}$ regimes, and is plotted
in Figure~\ref{fig:occ_rate_model_3_log}.

\paragraph{The rate of Earth analogs} The immediately arresting result
there is a ``bump'' in the apparent rate density at $r\a < 2r_\oplus$:
the true rate for singles is less than the inferred rate.  For the
case in which secondaries host as many (half as many) planets as
single stars, this means an overestimate of the occurrence
rate by $\approx 50\%$ ($\approx 25\%$).  The ``bump'' exists even for
the $Z_2/Z_0=0$ case as a $\approx 10\%$ effect.  The magnitude of
this error evidently depends strongly on the prevalence of planets
around secondaries.

\paragraph{Hot Jupiter occurrence rates} Taking
Figure~\ref{fig:occ_rate_model_3_log} and integrating the rate
density, we can compare the apparent hot Jupiter occurrence rate with
the true rate for singles.  In the most extreme case of $Z_2/Z_0=0$,
we find that $\Lambda_{{\rm HJ},0}/\Lambda_{\rm HJ,a} = 1.13$, where
\begin{equation}
    \Lambda_{\rm HJ,a} =
      \int_{8r_\oplus}^{\infty} \Gamma\a(r\a)\,{\rm d}r\a,
\end{equation}
and similarly for $\Lambda_{\rm HJ,0}$.
%occ_rate_vs_radius_model_3_withtext_Zsub2_0.00_rpu_22.5.pdf
If $Z_2/Z_0>0$, binarity affects the apparent hot Jupiter rate less:
when $Z_2/Z_0=0.5$, we find $\Lambda_{{\rm HJ},0}/\Lambda_{\rm HJ,a} =
1.06$.  We discuss this issue further in
Section~\ref{sec:further_models}.
